# Images, Surfaces, and Plotting

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(fields)
library(RColorBrewer)
```

In the course of spatial analysis of 2-d fields, one needs simple functions to create, plot, and smooth rectangular images. We present many of the `fields` functions for working with image data and simulatneously highlight the plotting capabilites of the package. By a surface or image object, we mean a list with components `x`, `y` and `z`, where `x` and `y` are equally spaced grids and `z` is a matrix of values.  This is the same format as used by the R functions `contour`, `image`, and `persp`. The `x` and `y` components index each pixel and also index the matrix, whereas the `z` component gives the pixel values.

## `as.image`

`as.image` creates an image object from irregular `x`, `y`, and `z` coordinates by discretizing the 2-d locations to a grid and then producing an image object with the `z` values in the correct cells. The basic idea is that each observation is identified with a grid box, and then the mean value for all observations is returned for that box. If no observations fall into a particular box, an `NA` is returned for that location. Missing values are handled deftly by the contouring and image functions, such as `image.plot`.
 
The required argument for `as.image` is `Z`, the values for the image and the most common optional argument is `x`, a 2 column matrix with the spatial locations of `Z`.  

In the following example we grid the `RMprecip` data and plot it. The default is a 64 by 64 grid using the ranges of the data. One can also specify alternative grids: the third panel in the next plot demonstrates a 25 by 40 point discretization.

```{r, results='hide', fig.width=11}
set.panel(1,3)
out.p <- as.image( RMprecip$y,  x= RMprecip$x)
image( out.p)
#compare to quilt.plot
quilt.plot(RMprecip$x, RMprecip$y)
grid<- list( x=seq( -111, -99,,25), y=seq( 35, 45,,40)) 
as.image( RMprecip$y, RMprecip$x, grid=grid)-> out.p
image(out.p)
```

Several other useful options are available. One can specify weights to use in finding a weighted mean for each box, or just give the number of grid points and use the ranges of the data for gridding. Also note that the list returned by `as.image` has components to indicate the counts in each bin and membership. 

&nbsp;

***

${\large{\mathbf{Basic \> Usage}}}$

> as.image(Z, ind=NULL, grid=NULL, x=NULL, nrow=64,   ncol=64,weights=NULL)    

${\large{\mathbf{Value}}}$

`as.image` returns a list: in addition to x,y, and z components, the list has the counts `N`, the indices of the nonmissing boxes `ind` and sum of the weights for each nonmissing box `weights`.
  
***
  
&nbsp;

## Plotting with `image`, `image.plot`

The function `image.plot` has the same functionality as `image` but adds a legend strip. It can be used to add a legend strip to an existing plot or create a new image and legend. Its chief benefit is that this function resizes the plot region automatically to make room for the legend strip. The argument `horizontal`, if true, will put the legend strip under the plot. One effective graphical plot is a panel of images on a similar scale and plot region size but with only one legend strip. `image.plot` makes this fairly easy to do.   

The function `add.image` is useful for adding an image to an existing plot. The user gives the matrix of intensities of the image as `z`, as well as simple arguments that control the location and size. `xpos` and `ypos` give the position of the image, and `adj.x` and `adj.y` determine whether the image is centered on that location (0.5,0.5), to the bottom right (1,1), Finally, we can change the width and height of the image as a fraction of the plotting region with `image.width` and `image.height`. If `NULL`, height is scaled to make image pixels square.

To lead by example, we plot the `RMprecip` data at several discrtetizations. We see that indeed only `image.plot` includes a colorbar, and the color palettes are quite different. Then we use `add.image` to put John on the Four Corners for fun!

```{r, results='hide', fig.width=12}
set.panel( 1,3)
par( pty="s")   # square plotting regions 
image.plot( as.image( RMprecip$y, x = RMprecip$x), col=heat.colors(12)); US( add=TRUE, col="black", lwd=2)
image(  as.image( RMprecip$y, x = RMprecip$x, ncol=32,nrow =32)); US( add=TRUE, col="black", lwd=2)
image(as.image( RMprecip$y, x = RMprecip$x, ncol=16, nrow=16)); US( add=TRUE, col="black", lwd=2)
data(lennon)
add.image( -109,37, lennon , col=grey( (0:256)/256))
```

&nbsp;

***

${\large{\mathbf{Basic \> Usage}}}$

> image.plot(image, col = tim.colors(nlevel))  
> add.image(xpos, ypos, z, adj.x = 0.5, adj.y = 0.5, image.width = 0.15, image.height = NULL)  

***
  
&nbsp;

<Here is an example using the full functionality of `image.plot`. >

```{r,  include=FALSE, eval=FALSE}
x<- 1:10
y<- 1:15
z<- outer( x,y,"+")
image.plot(x,y,z)
# or
obj<- list( x=x,y=y,z=z)
image.plot(obj)
# adding breaks and distinct colors for intervals of z
# with and without lab.breaks
brk<- quantile( c(z))
colorTable<- designer.colors(4, c( "blue","purple", "red") )
image.plot(x,y,z, breaks=brk, col=colorTable, lab.breaks=names(brk))
# annotate legend strip at break values and add a label
# compare to
zp <-quantile(c(z), c( .05, .1,.5, .9,.95))
image.plot(x,y,z, axis.args=list( at=zp, labels=names(zp) ) )
```

***

## `grid.list`, `make.surface.grid`, and `as.surface`


`grid.list` refers to an x, y list format for a grid. `make.surface.grid` (a wrapper around the R base function `expand.grid`) is a useful function that expands the grid from the `grid.list` into a full set of locations.

`as.surface` is a similar function to `as.image`. This function was written to simply to go back and forth between a matrix of gridded values and the stacked vector obtained by stacking columns. The input `obj` can be a `grid.list` of x and y coordinates, or a matrix of the grid points produced by `make.surface.grid`. `z` denotes the values of the function at the grid points, and finally `order.variables` (either `"xy"` or `"yx"`) specifies how the x and y variables used to evaluate the function are matched with the x and y grids in the surface object. (Note: to convert irregular 2-d data to a surface object where there are missing cells, see the function `as.image`.)


&nbsp;

***

${\large{\mathbf{Basic \> Usage}}}$

> make.surface.grid(grid.list)  
> as.surface(obj, z, order.variables="xy")    

${\large{\mathbf{Value}}}$

The result will be a list with x, y, and z components suitable for plotting with functions such as `persp`, `image`, `surface`, `contour`, `drape.plot`.

`make.surface.grid` returns a full set of grid locations as a matrix, and `as.surface` returns a list of class `surface`.

***

&nbsp;

```{r}
grid.list <- list( x= -111:-99, y=35:45)
xg <- make.surface.grid(grid.list)
str(grid.list)
str(xg)
```

A typical operation is to go from a `grid.list` to the set of grid locations, evaluate a function at these locations, and then reformat this as an image/surface for plotting. Here is how to do this cleanly:

```{r}
fit <- spatialProcess(RMprecip$x, RMprecip$y, aRange=20)
grid.list <- list( x= -111:-99, y=35:45)
xg<- make.surface.grid(grid.list)
look <- predict(fit, xg)
look.surface <- as.surface(xg, look)
```

The next example gives the user an idea of the comparison of the plotting functions in `fields` or the standard libraries that are relevant to images, surfaces, grids, and object in classes defined in `fields`. 

```{r}
image(look.surface)
```

Using `image.plot` in `fields` adds a colorbar and a different default color scale. 

```{r}
image.plot(look.surface)
```

Here is the standard R function `contour`.  

```{r}
contour(look.surface)
```

Using `surface` combines `image.plot` with `contour` as its default.

```{r}
surface(look.surface)
```

There is a graphics function `persp` which can produce 3-d plots. `drape.plot` is a wrapper function in `fields` that automatically adds color based on how it's done in `image.plot`/`surface`.  

```{r, results='hide', fig.width=12}
set.panel(1,2)
persp(look.surface)
drape.plot(look.surface)
```

In summary, we fit a `spatialProcess` object from our data, created a "grid.list", constructed a full set of gridded locations using the "grid.list," found Kriging estimates at the gridded locations, converted back to a "grid.list", and then plotted with `surface`. Compare this with the convenient `surface.mKrig`.

```{r}
fit <- spatialProcess(RMprecip$x, RMprecip$y, aRange=20)
surface(fit)
```


## `surface`

We saw in the last example that we can use `surface` to visualize both `surface` objects and also `spatialProcess`/`mKrig`/`Krig` objects. It is a great tool to be able to fit a class object and quickly plot predictions in a few short lines of code.

This `surface` function is essentially a combination of `predictSurface` and `plot.surface`. It may not always give a great rendition but is easy to use for checking the fitted surface. The default of `extrap=F` is designed to discourage looking at the estimated surface outside the range of the observations. Note that any `Z` covariates in the the model will be dropped and only the spatial part of the model will be evaluated.

&nbsp;

***

${\large{\mathbf{Basic \> Usage}}}$

> surface( object, grid.list = NULL, extrap = FALSE, type="C")  

***

&nbsp;

The `type` argument has the following options: `"p"` gives `persp`, `"c"` gives `contour` with legend strip (`image.plot`), and `"C"` gives `image plot` with contours overlaid.


```{r}
quilt.plot(ChicagoO3$x, ChicagoO3$y,col=terrain.colors(50))
```

```{r, results='hide', fig.width=11}
fit<- Krig(ChicagoO3$x,ChicagoO3$y, aRange=30) 
set.panel(1,2)
surface(fit, type="C", nx=128, ny=128,col=terrain.colors(50))
surface(fit, type="p", nx=56, ny=56, extrap = TRUE, border=NA, aRange=-8, phi=20,col=terrain.colors(50))
```